Maximum Entropy Spectral Analysis: a case study

### 2021-04-16 paper/mesa.tex
conclusions and more references

\section{Final remarks and future prospects} \label{sec:conclusion}
\sschmidt{Check this section}
We presented a case study of the application of Maximum Entropy principle in the realm of spectral estimation: it provides a closed form estimation of the power spectral density, bridging the spectral analysis with the study of autoregressive processes. The method has the towfold advantage of avoiding any unjustified assumption on the observed time series and of providing a simple context for forecasting the time series. We provided a ready to use python code, called \texttt{memspectrum} and we performed numerical studies of its performance.
We investigated the use of different loss functions for the purpose of correctly estimating the autoregressive order for the estimation and we found that different loss functions are required for different situations.
Finally, we performed a comparison between PSDs computed with MESA and standard Welch's algorithm and we applied our method to several time series of interest in physics.

As outlined in sec.~\ref{sec:Welch_comparison}, the MESA provides PSD estimates with less noise than the standard Welch algorithm, thus making it a good alternative for it. The use of MESA is particularly useful for short time series samples, where Welch method is outperformed in accuracy and confidence.
It has to be said that our method can be quite computationally expensive, as compared to standard Welch's algorithm. However, for short time series, the computational cost is still managable.

Besides providing an alternative to standard methods of spectral estimate, MESA opens up several oppurtunities in the context of time series analysis.

A first application is based on the observation that meaningful PSD estimations can be extracted from short time series (unlike the standard approach).
This can make the approach particularly conveniente in the context of Bayesian parameter estimation, whenever the likelihood of the model for a time series takes the form
\begin{equation}
	\mathcal{L}(d_t | \theta) \propto exp\left[ -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2}{S(f)} \right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes Fourier transform and $x_t(\theta)$ is a prediction for a signal hidden in the observed time series $d(t)$, dependent on some parameter $\theta$.
This likelihood corresponds to looking for a signal $x_t$ into some (possibly very) noisy data.
Usually the PSD is estimated offline on a large batch of data, that might also contain the signal to analyse and in general it might not reflect the structure of noise in the time slice to analyse.
MESA can offer a way out from this problem: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residual of noise subtracted with the signal.
%To compute the posterior distribution for the parameters the knowledge of prior distributions and Likelihood for noise is required. A full description for noise can be obtained in Fourier Domain by the choice of the Whittle Likelihood \eqref{eq:Whittle}, that is completely determined by the knowledge of the PSD for noise. While traditional methods might be instable with respect to the hypothesis (such as Welch) or require a Large number of prior assumptions, MESA represents a stable and parsimonious way to characterize noise and compute posterior distributions.

Furthermore, MESA provides a simple, but robust and quite accurate in the short term, predictor for the time series. This fact is remarkable and can be used in time series analysis for several purposes.
As already discussed in sec~\ref{sec:LIGO_forecasting}, an anomaly detection pipeline can be easily built using the forecasts of MESA: the predictions can form a baseline to compare the actual observations with. Whenever the observed data are outside the expectations, an anomaly detection can be claimed.
Of course such predictions can be done with a more accurate (perhaps nonlinear model); however MESA has the advantage of being simple and fast to train, while providing good (short term) predictions.

\sschmidt{Other stuff with forecasting?}

We believe that Maximum Entropy Spectral Analysis, despite being an ``old" theory, has still the potential to become the standard in future time series analysis, due to its theoretical elegance and to its reliable performance in a vast range of application.

>>>>

\section{Final remarks and future prospects} \label{sec:conclusion}
In this work, we presented a case study of the application of Maximum Entropy principle in the realm of spectral estimation. Albeit the methodology hereby presented is grounded on solid theoretical foundations and its merits are widely recognised, Maximum Entropy methods have yet to be adopted routinely in the study of problems related to time series. The superior nature of maximum entropy methods and in particular of Burg's method is exemplified by the closed form estimate of the power spectral density and by the theoretical bridge it lies between spectral analysis and autoregressive processes. Moreover, the method presents, in our view, two main advantages when compared with more traditional ones; first there is no need to choose an arbitrary window function to correct the data and second it provides as straightforward way to compute predictions given past observations. Accompanying this work, we provide a publicly available \texttt{Python} implementation, called \texttt{memspectrum}, that we used to perform numerical studies of its behaviour and to  demonstrate its reliability. 
Since the order of the autoregressive process is not determined by the theory, we opted for an in-depth investigation of several proposals in the literature and found that different loss functions are required for different situations, with the FPE loss function being the most indicated to deal with gravitational wave data. Along these lines, we directly compared the PSDs computed with MESA with the canonical Welch's algorithm. As outlined in sec.~\ref{sec:Welch_comparison}, MESA provides PSD estimates with smaller variance than Welch algorithm. The use of MESA is particularly useful for short time series samples, where Welch method is outperformed in accuracy and confidence. 
This observation suggests a promising avenue to pursue in future developments: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD, hence MESA could be employed for a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalising over the detector noise PSD. Whenever the likelihood of the model for a time series takes the form
\begin{equation}
	\mathcal{L}(d_t | \theta) \propto exp\left[ -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2}{S(f)} \right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$. The typical approach is to estimate the PSD offline on a large batch of data and off-source (see~\cite{lalinference}), that might not reflect the structure of noise in the time slice to analyse. MESA can offer a way out from this problem: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. Preliminary studies suggest that this is possible~\cite{martini_thesis}.

Furthermore, MESA provides a simple, but robust and quite accurate in the short term, predictor for the time series. This fact is remarkable and can be used in time series analysis for several purposes. As discussed in Sec.~\ref{sec:LIGO_forecasting}, an anomaly detection pipeline can be easily built using the forecasts of MESA: the predictions can form a baseline to compare the actual observations with. Whenever the observed data are outside the expectations, an anomaly detection can be claimed. Of course such predictions can be done with a more accurate (perhaps nonlinear model); however MESA has the advantage of being simple and fast to train, while providing good (short term) predictions. At the same time, several instruments present gaps in their data stream, for instance LISA is expected to show such gaps~(e.g. \cite{lisa_gaps} and references therein), MESA forecasting capabilities could be used to fill those gaps with predicted data from past observations.

In conclusion, we reiterate that MESA is a theoretically sound, computationally feasible and reliable way of studying the properties of stochastic processes and we hope that the investigations presented in this work will further stimulate developments and applications of this method.
